---
title: "eCSM projections"
author: "Sigurd Maurud"
date: "2025-05-10"
output:
  pdf_document:
    toc: yes
    toc_depth: 5
    latex_engine: xelatex
    keep_tex: yes
  word_document:
    toc: yes
    toc_depth: '4'
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Integrate Python in R Markdown 

```{r}
library(reticulate)
use_python("C:/Program Files/Python312/python.exe", required = TRUE)
```

# 1. Text preparations and wrangling

Load data

```{python}
from docx import Document

# Function to read text files from file path
def load_transcripts(path):
  transcript = Document(path) # create transcript object 
  text = [] # initalise list for storing text
  for paragraph in transcript.paragraphs: # iterate over paragraphs
    if paragraph.text.strip(): # Omit empty paragraphs
      text.append(paragraph.text) # Append text to intialised list
  return ' '.join(text) # Return the combined text and paragraphs

# Define file path for all 17 interview transcripts
paths = [
    "Interviews/Raw files/Transcript 1.docx",
    "Interviews/Raw files/Transcript 2.docx",
    "Interviews/Raw files/Transcript 3.docx",
    "Interviews/Raw files/Transcript 4.docx",
    "Interviews/Raw files/Transcript 5.docx",
    "Interviews/Raw files/Transcript 6.docx",
    "Interviews/Raw files/Transcript 7.docx",
    "Interviews/Raw files/Transcript 8.docx",
    "Interviews/Raw files/Transcript 9.docx",
    "Interviews/Raw files/Transcript 10.docx",
    "Interviews/Raw files/Transcript 11.docx",
    "Interviews/Raw files/Transcript 12.docx",
    "Interviews/Raw files/Transcript 13.docx",
    "Interviews/Raw files/Transcript 15.docx",
    "Interviews/Raw files/Transcript 17.docx",
    "Interviews/Raw files/Transcript 18.docx",
    "Interviews/Raw files/Transcript 20.docx"]

# Load transcripts and store them in a list
transcripts = [] # initialise list for storing transcripts
for path in paths: # iterate over file paths
  transcript_text = load_transcripts(path)
  transcripts.append(transcript_text)

print(f'Number of transcripts:\n{len(transcripts)}')
```

Import JSON list of defined corrections and list of stopwords (secure environment allows limited internet access) 

```{python}
import json
import pickle

with open('corrections.json', 'r', encoding='utf-8') as file:
  corrections = json.load(file)

with open('stopwords.pkl', 'rb') as f:
  stopwords = pickle.load(f)
```

Tokenise text, remove special characters, lowercase letters, correct misspellings and remove stopwords. 

```{python}
import nltk
nltk.data.path.append('nltk_data') # Local import of NLTK functionality
from nltk.tokenize import sent_tokenize, word_tokenize
import re
from collections import Counter

# 1. Tokenise text into sentence tokens and word tokens 
sent_tokens = []
word_tokens = []

# Sentence tokens 
for transcript in transcripts:
  sentences = sent_tokenize(transcript, language='norwegian')
  sent_tokens.extend(sentences)

# Word tokens 
for transcript in transcripts:
  words = word_tokenize(transcript, language='norwegian')
  word_tokens.extend(words)

# Total vocabulary
vocabulary = Counter(word_tokens)

# 2. Function to remove special characters

def remove_special_characters(text, remove_digits=False):
# Regular expression. ^ equals negation: "all that is not" letters and whitespace (\s)
    pattern = r'[^a-zA-ZæøåÆØÅ0-9\s]' if not remove_digits else r'[^a-zA-ZæøåÆØÅ\s]'  
# Replace all special characters and numbers that match pattern with empty string
    text = re.sub(pattern, '', text)
    return text
  
# Remove special characters
clean_word_tokens = [remove_special_characters(token, remove_digits=True) for 
token in word_tokens]
clean_sent_tokens = [remove_special_characters(token, remove_digits=True) for 
token in sent_tokens]

# 3. Lowercase text

def lowercase(tokens):
  return [token.lower() for token in tokens]

clean_word_tokens = lowercase(clean_word_tokens)
clean_sent_tokens = lowercase(clean_sent_tokens)

# 4. Define function for correcting misspellings

def correct_spelling(text, correction_map):
  # Tokenise sentence tokens into words
    words = nltk.word_tokenize(text.lower(), language='norwegian')
    corrected_words = []# Initialise list for corrected text

    for word in words:
      # Replace misspellings with corrections
        corrected_word = correction_map.get(word, word) 
        if ' ' in corrected_word:
          # Split corrections consisting of several words into independent words 
            corrected_words.extend(corrected_word.split(' ')) 
        else:
            corrected_words.append(corrected_word)
    
    return ' '.join(corrected_words)

# Correct misspellings
clean_sent_tokens = [correct_spelling(sentence, corrections) for 
sentence in clean_sent_tokens]
clean_word_tokens = [correct_spelling(word, corrections) for 
word in clean_word_tokens]

# 5. Define function to remove stopwords
def remove_stopwords(sentence):
    word_list = sentence.split()
    # Filter stopwords
    no_stopwords = [word for word in word_list if word not in stopwords] 
    return ' '.join(no_stopwords)

# Remove stopwords
clean_sent_tokens = [remove_stopwords(sentence) for sentence in clean_sent_tokens]
no_stopword_tokens = [word_tokenize(sentence) for sentence in clean_sent_tokens]
clean_word_tokens = [word for sublist in no_stopword_tokens for word in sublist]
clean_vocabulary = Counter(clean_word_tokens)

# 6. Inspect
print(f'Total number of sentences:\n {len(sent_tokens)}')
print(f'Total number of words:\n {len(word_tokens)}')
print(f'Total number of independent words:\n {len(vocabulary)}')
print(f'Total number of clean sentences:\n {len(clean_sent_tokens)}')
print(f'Total number of words with stopwords omitted:\n{len(clean_word_tokens)}')
print(f'Total number of independent words with stopwords omitted:\n{len(clean_vocabulary)}')
```

Convert clean sentence tokens into data frame

```{python}
import pandas as pd

sents_df = pd.DataFrame({'Sentence': clean_sent_tokens})
print('Number of rows and columns in dataframe: \n', sents_df.shape, '\n')
```

Vectorise data frame into frequency-based Bag of Words (BoW)

```{python}
from sklearn.feature_extraction.text import CountVectorizer
import numpy as np

# Create Count vectoriser object that considers unigrams and bigrams
cv = CountVectorizer(ngram_range=(1,2))
# Train BoW on unigram and bigram vocabulary and create sparse matrix
bow = cv.fit_transform(sents_df['Sentence'])
# Convert BoW to numpy array
bow_matrix = bow.toarray()
# Get vocabulary from BoW
bow_vocab = cv.get_feature_names_out()
# Convert to Data Frame for inspection
bow_df = pd.DataFrame(np.round(bow_matrix, 2), columns=bow_vocab)
print(f'Number of rows and columns in data frame:\n{bow_df.shape}')
print('First five rows of data frame: ') 
print(bow_df.head)
```

# 2. Topic model - Non-negative matrix factorisation 

Compute coherence $C_v$ for different number of topics

```{python}
from sklearn.decomposition import NMF
from gensim.corpora.dictionary import Dictionary
from gensim.models.coherencemodel import CoherenceModel

# Define different numbers of topics to assess
num_topics = [2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28, 30]

# Initialise list for coherence scores
coherence_cv = []

# Set up Gensim dictionary of vocabulary for computing coherence 
gensimDictionary = Dictionary(no_stopword_tokens)
# Create Gensim BoW corpus for computing coherence 
corpus = [gensimDictionary.doc2bow(doc) for doc in no_stopword_tokens]

for n_topics in num_topics:
  # Train NMF with specified parameters
  nmf = NMF(n_components=n_topics, solver='cd', max_iter=1000, random_state=42)
  nmf.fit(bow_matrix)
  topic_terms = nmf.components_ # Extract topics
  
  # Extract top 10 topic words 
  topics = []
  for topic_idx, topic in enumerate(topic_terms):
    # last 10 indices in descending order
    topic_words = [bow_vocab[i] for i in topic.argsort()[:-10 - 1:-1]] 
    topics.append(topic_words)
  # Compute coherence for model topics based on top 10 words 
  coherence_model = CoherenceModel(topics=topics, texts=no_stopword_tokens, 
  dictionary=gensimDictionary, coherence='c_v')
  # Store results 
  coherence_cv.append(coherence_model.get_coherence())
```

Assess coherence for different number of topics

```{python}
list = num_topics
for num_topics, coherence in zip(list,coherence_cv):
    print(f"Number of topics: {num_topics}, Coherence score C_v: {coherence}")
```

Train NMF with 28 topics

```{python}
# Define number of topics
n_topics = 28
# Train NMF and define V, W, and H matrices
V = bow_matrix
nmf = NMF(n_components=n_topics, solver='cd', max_iter=1000, random_state=42)
W = nmf.fit_transform(V)
H = nmf.components_ 
feature_names = bow_vocab

# Build data frame with top 10 words for each topic
topic_key_terms_idx = np.argsort(-np.absolute(H), axis=1)[:, :10]
# Sort top 10 indices for each topic in descending weights 
topic_keyterms = feature_names[topic_key_terms_idx] # Name each indices by their word
# Combine keyterms for each topic in a string
topics = [', '.join(topic) for topic in topic_keyterms] 
# Create data frame with indexation for each topic
Topics_df = pd.DataFrame(
  topics, columns=['Terms per Topic'], 
  index=['Topic'+str(i) for i in range(1, n_topics+1)]
  ) 
```

Visualise top 10 words for each topic

```{r}
# Export data frame to R
Topics <- py_to_r(py$Topics_df)
# Ensure data frame is a data frame
Topics <- as.data.frame(Topics)
# Print
knitr::kable(Topics)
```

Export Topics data frame and sentence tokens for semantic assessment of each topic

```{python}
with open('sent_tokens.pkl', 'wb') as f:
  pickle.dump(sent_tokens, f)

Topics_df.to_csv("Topics.csv", index=True, header=True)
```

Transfer matrix W to R

```{r, fig.width=12, fig.height=8, message=FALSE}
library(dplyr)
# Transfer W matrix
W1 <- py_to_r(py$W)
# Transform W1 to dataframe
W1 <- as.data.frame(W1)
# Name topics based on identified content 
W1 <- W1 %>%
  rename_at(vars(1:28), ~ c("1. Information needs in current offer",
                            "2. Living with IBD",
                            "3. Digital content and services",
                            "4. Bad periods and concerns", 
                            "5. Disease understanding",
                            "6. IBD help line",
                            "7. Existing sources of information",
                            "8. Health service communication and follow-up",
                            "9. Health information: availability and quality",
                            "10. Possibilities in digitisation",
                            "11. Medicine information",
                            "12. Abstinence and healthy habits",
                            "13. Nutrition", 
                            "14. Information access before and now",
                            "15. Symptoms: frequency and intensity",
                            "16. Primary care",
                            "17. Manage daily living",
                            "18. Wishes for information and services",
                            "19. Iron deficiency",
                            "20. Information exchange with HCP",
                            "21. Symptoms and affiliated ailments",
                            "22. Effect of medicines",
                            "23. Contact with the health service",
                            "24. Identifying the disease",
                            "25. Diet",
                            "26. Internet and search engines",
                            "27. Challenges in information 
                            retrieval and needs assessment",
                            "28. Digitisation as driver for 
                            active patient involvement"))
```

# 3. Project free-text responses to topic model

Load data

```{r}
HL_DHL <- read.csv2('HLDHL.data.csv')
# Combine the three Brief IPQ free-text responses into one variable 
BIPQ9 <- paste(HL_DHL$BIPQ9_1, HL_DHL$BIPQ9_2, HL_DHL$BIPQ9_3, sep = " ")
# Transfer to Python for text analysis
py$BIPQ9 <- BIPQ9
```

Define additional corrections for identified words in free-text responses

```{python}
corrections_2 = {
  "abbruk": "antibiotika bruk",
  "fam": "familie",
  "stresstraumalivssituasjon": "stress traume livssituasjon",
  "arvdisponert": "arv disponert",
  "virusinfeksjonbakterie": "virus infeksjon bakterie",
  "seksueltvoldelig": "seksuelt voldelig",
  "depresjonangst": "depresjon angst",
  "bechterews": "bekhterevs",
  "arvgenetisk": "arv genetisk",
  "laaang": "lang",
  "genetikkarv": "genetikk arv", 
  "arvgener": "arv gener",
  "tetracyckin": "tetracyklin",
  "sovn": "søvn",
  "overthinking": "overtenking",
  "mtp": "med tanke på", 
  "osv": "og så videre",
  "sundt": "sunt",
  "røket": "røyket",
  "mnd": "måned",
  "giftforgiftet": "gift forgiftet",
  "arveligfarmor": "arvelig farmor",
  "stressangst": "stress angst",
  "feks": "for eksempel",
  "bectrews": "bekhterevs",
  "imunsystem": "immunsystem",
  "uc": "ulcerøs kolitt",
  "bakteriematmijlø": "bakterie mat miljø",
  "emotionell": "emosjonell",
  "adhdautisme": "adhd autisme",
  "milj": "miljø",
  "mtp": "med tanke på",
  "hyper prosessert": "hyperprosessert",
  "hemorider": "hemoroider",
  "livstill": "livsstil",
  "uflaksgener": "uflaks gener",
  "forstyrretpåvirket": "forstyrret påvirket",
  "ikkr": "ikke",
  "arvmiljøtilfeldig": "arv miljø tilfeldig",
  "svingingerpubertet": "svingninger pubertet",
  "svakedårlige": "svake dårlige",
  "arveliggenetisk": "arvelig genetisk",
  "stressjobbensomhet": "stress jobb ensomhet",
  "gut": "tarm",
  "bacteria": "bakterie",
  "distribution": "distribusjon",
  "genetic": "genetisk",
  "ultraprossesert": "ultraprosessert",
  "div": "diverse",
  "stresstraumer": "stress traumer",
  "kostholdlivsstil": "kosthold livsstil",
  "el": "eller",
  "rsaken": "årsaken"}
```

Predict free-text responses to NMF topics

```{python, message=FALSE, warning=FALSE}
# Word tokens of free-text responses
words_BIPQ9 = []
for obs in BIPQ9:
  words = word_tokenize(obs, language='norwegian')
  words_BIPQ9.extend(words)

vocab_BIPQ9 = Counter(words_BIPQ9)

# Preprocess and wrangle text from free-text responses 
clean_BIPQ9 = [remove_special_characters(token, remove_digits=True) for token in BIPQ9]
clean_BIPQ9 = lowercase(clean_BIPQ9)
clean_BIPQ9 = [correct_spelling(sentence, corrections) for sentence in clean_BIPQ9]
clean_BIPQ9 = [correct_spelling(sentence, corrections_2) for sentence in clean_BIPQ9]
clean_BIPQ9 = [remove_stopwords(sentence) for sentence in clean_BIPQ9]

# Word tokens of clean free-text responses

words_clean_BIPQ9 = []
for obs in clean_BIPQ9:
  words = word_tokenize(obs, language='norwegian')
  words_clean_BIPQ9.extend(words)

vocab_clean_BIPQ9 = Counter(words_clean_BIPQ9)
# Inspect
print(f'Total number of observations:\n {len(BIPQ9)}')
print(f'Total number of words:\n {len(words_BIPQ9)}')
print(f'Total number of independent words:\n {len(vocab_BIPQ9)}')
print(f'Total number of words with stopwords omitted:\n{len(words_clean_BIPQ9)}')
print(f'Total number of independent words with stopwords omitted:\n{len(vocab_clean_BIPQ9)}')
```

Project topics in free-text responses
```{python}
BIPQ9_bow = cv.transform(clean_BIPQ9)
BIPQ9_W = nmf.transform(BIPQ9_bow)
```

Export W matrix to R

```{r}
W2 <- py_to_r(py$BIPQ9_W)
```

Compare mean topic presence in W-matrices

```{r, warning=FALSE, message=FALSE}
W.comparance <- data.frame(
  mean_W1 = round(colMeans(W1), digits = 4),
  topic_presence_W1 = round((colSums(W1 > 0.01)/nrow(W1))*100, digits = 2),
  mean_W2 = round(colMeans(W2), digits = 4),
  topic_presence_W2 = round((colSums(W2 > 0.01)/nrow(W2))*100, digits = 2)
)
knitr::kable(W.comparance)
```

```{r}
relevant.W2 <- (W.comparance$mean_W2 > W.comparance$mean_W1) &
  (W.comparance$topic_presence_W2 > W.comparance$topic_presence_W1)
knitr::kable(W.comparance[relevant.W2 == TRUE, ])
```

Set topic names for W2

```{r}
W2_nonames <- W2 # Save a copy of W2 without topic names for later processing
W2 <- as.data.frame(W2)
colnames(W2) <- colnames(W1)
```

# 4. Principal component analysis

Perform PCA. Omit scaling as topics are weighted on the same scale and represent relative relationships

```{r}
pr.W <- prcomp(W2, center = TRUE, scale. = FALSE)
summary(pr.W) # Print output
```

Plot proportion of variance explained and cumulative proportion of variance explained by each component

```{r, fig.width=6, fig.height=4}
par(mfrow = c(1, 2))
plot(summary(pr.W)$importance[2, ], xlab = "Principal component", 
     ylab = "Proportion of variance explained", ylim = c(0,1), 
     type = "b")
plot(summary(pr.W)$importance[3, ], xlab = "Principal component", 
     ylab = "Cumulative proportion of variance explained", ylim = c(0,1), 
     type = "b")
```

Extract the first six components from PCA to account for 90 % variance

```{r}
w.pcs <- pr.W$x[, 1:6]
```

Inspect the contribution of each topic into the components

```{r}
knitr::kable(round(pr.W$rotation[, 1:6], digits = 3))
```

# 5. Hierarchical clustering of first 6 principal components

Assess agglomerative coefficient (AC) for different combinations of distances and linkages

```{r, message=FALSE, warning=FALSE}
library(cluster)
library(proxy)
set.seed(43) 

# Define distances and linkages
distances <- c("euclidean", "maximum", "manhattan", "canberra", "cosine")
linkages <- c("average", "single", "complete", "ward")

# Function for AC-evaluation
clustEv <- function(dist.m, link.m) {
  copy <- w.pcs # use copy of data to avoid interference
  dist.m <- if (dist.m == "cosine"){ # stats::dist() does not support cosine
    as.dist(proxy::dist(copy, method = "cosine"))
  } else {
    stats::dist(copy, method = dist.m) # return to integrated stats()-function
  }
  res.agnes <- agnes(dist.m, method = link.m)
  return(res.agnes$ac)
}


res <- expand.grid(distance = distances, linkage = linkages) %>%
  rowwise() %>%
  mutate(ac = clustEv(distance, linkage)) %>% 
  ungroup()

# View results
knitr::kable(res)
```

Continue with Cosine distance and complete linkage

```{r, fig.width=6, fig.height=4}
dist.cos <- proxy::dist(w.pcs, method = "cosine")
cos.com <- hclust(dist.cos, method = "complete")
plot(cos.com, hang = -1, cex = 0.5, main = "Cosine complete")
```

Dendrogram indicates three distinct branches. Assess Silhouette coefficient for cluster solutions.

```{r, fig.width=6, fig.height=4}
sil.k <- list()
for (k in 2:6) {
  clusters <- cutree(cos.com, k = k)
  # Silhouette
  sil.k[[k]] <- mean(cluster::silhouette(clusters, dist.cos)[, 3])
}
silhouette <- unlist(sil.k)
names(silhouette) <- c(2:6)
plot(2:6, silhouette, main = "Silhouette coefficient",
     xlab = "Number of clusters", ylab = "Silhouette", type = "b")
# Assess mean coefficients with different number of clusters 
print(silhouette)
```

Assess stability of cluster solutions through bootstrapped Jaccard similarity index(JSI)

```{r, warning=FALSE}
library(fpc)
set.seed(43)
# Define function for fpc::clusterboot()
CosCom <- function(x, k) {
  dist.matrix <- dist(x, method = "cosine")
  dist.m <- as.matrix(dist.matrix)
  hc <- hclust(dist.matrix, method = "complete")
  clusters <- cutree(hc, k = k)
  clusterlist <- lapply(1:k, function(i) clusters == i)
  list(
    result = hc,
    nc = k,
    clusterlist = clusterlist,
    partition = clusters,
    clustermethod = "hierarchical"
  )
}

# Compare mean JSI over bootstrap samples
cut3 <- clusterboot(w.pcs, B = 1000, bootmethod = "boot",
                      clustermethod = CosCom, k = 3, count = FALSE)
cut4 <- clusterboot(w.pcs, B = 1000, bootmethod = "boot",
                      clustermethod = CosCom, k = 4, count = FALSE)
cut5 <- clusterboot(w.pcs, B = 1000, bootmethod = "boot",
                      clustermethod = CosCom, k = 5, count = FALSE)
cut6 <- clusterboot(w.pcs, B = 1000, bootmethod = "boot",
                      clustermethod = CosCom, k = 6, count = FALSE)

# View results
cut3$bootmean
cut4$bootmean
cut5$bootmean
cut6$bootmean
```

Assess cluster distribution for different solutions

```{r}
cut3 <- cutree(cos.com, k = 3)
cut4 <- cutree(cos.com, k = 4)
cut5 <- cutree(cos.com, k = 5)
cut6 <- cutree(cos.com, k = 6)
W2clusters <- mutate(as.data.frame(W2_nonames), Cluster3 = as.factor(cut3), 
              Cluster4 = as.factor(cut4), Cluster5 = as.factor(cut5), 
              Cluster6 = as.factor(cut6))
summary(W2clusters$Cluster3)
summary(W2clusters$Cluster4)
summary(W2clusters$Cluster5)
summary(W2clusters$Cluster6)
```


***Conclusion:*** *Silhouette coefficient and JSI demonstrate high values​for k = 3 up to k = 6, but at k higher than 3, small clusters are generated, posing risk of reduced interpretative value. A 3-cluster solution is indicated by the dendrogram and provides the most balanced distribution.*

Merge selected clusters with corresponding observations in data

```{r}
W2c <- mutate(as.data.frame(W2_nonames), Cluster = as.factor(cut3))
```

Assess cluster profiles for topics

```{r}
topic.profiles.1 <- W2c %>%
  group_by(Cluster) %>%
  summarise('2. Living with IBD' = round(mean(V2), digits = 3), 
            '2. SD' = round(sd(V2), digits = 3),
            '2. Topic presence' = round(mean(V2 > 0) * 100, digits = 3),
            '4. Bad periods and concerns' = round(mean(V4), digits = 3), 
            '4. SD' = round(sd(V4), digits = 3),
            '4. Topic presence' = round(mean(V4 > 0) * 100, digits = 3),
            '21. Symptoms and affiliated ailments' = round(mean(V21), digits = 3), 
            '21. SD' = round(sd(V21), digits = 3),
            '21. Topic presence' = round(mean(V21 > 0) * 100, digits = 3),
            '25. Diet' = round(mean(V25), digits = 3), 
            '25. SD' = round(sd(V25), digits = 3),
            '25. Topic presence' = round(mean(V25) * 100, digits = 3))
# Transpose table
topic.profiles.1 <- as.data.frame(t(topic.profiles.1))
# Round values
topic.profiles.2 <- as.data.frame(lapply(topic.profiles.1, 
                                         function(x) round(as.numeric(x), 5)))
# Give columns names
colnames(topic.profiles.2) <- c('Cluster 1', 'Cluster 2', 'Cluster 3')
rownames(topic.profiles.2) <- rownames(topic.profiles.1)
# Print topic profiles
knitr::kable(topic.profiles.2[-1,])
```



## Cluster profiles for metadata

Compute EQ-5D-5L according to Norwegian value set

```{r, warning=FALSE, message=FALSE}
library(eq5d)
# Define dimensions according to function
scores.df <- data.frame(
  MO = EQdimensions$EQ5D_mobility, 
  SC = EQdimensions$EQ5D_self_care, 
  UA = EQdimensions$EQ5D_usual_activities, 
  PD = EQdimensions$EQ5D_pain_discomfort, 
  AD = EQdimensions$EQ5D_anxiety_depression)
# Compute index
eq5d5l <- eq5d(scores.df, country = "Norway", version = "5L", 
               type = "VT", ignore.invalid = TRUE)
```

Cluster profiles for continuous variables

```{r}
# Get continuous variables from HL/DHL data
HL.con <- HL_DHL[, c(2, 8, 19, 25:27, 31:46)]
real_HL <- read.csv2("N:/durable/OUS-GHsurvey/Sigurd - surveydata/Analyse/HL_DHL_data.csv")
BIPQ18 <- real_HL[, 55:62]
# merge data into one dataframe
HL.con <- mutate(HL.con, EQindex = eq5d5l, Cluster = W2c$Cluster, Consequences = BIPQ18$BIPQ1, Timeline = BIPQ18$BIPQ2, Personal_control = BIPQ18$BIPQ3, Treatment_control = BIPQ18$BIPQ4, Identity = BIPQ18$BIPQ5, Concern = BIPQ18$BIPQ6, Comprehensibility = BIPQ18$BIPQ7, Emotions = BIPQ18$BIPQ8)
# Get summaries 
con.profiles <- HL.con %>% 
  group_by(Cluster) %>%
  summarise('Age' = mean(age), 
            'SD Age' = sd(age), 
            'Disease duration' = mean(disease_duration), 
            'SD disease duration' = sd(disease_duration), 
            'EQ VAS' = mean(EQ5Dvas1, na.rm = T), 
            'SD EQ VAS' = sd(EQ5Dvas1, na.rm = T),
            'EQ index' = mean(EQindex, na.rm = T), 
            'SD EQ index' = sd(EQindex, na.rm = T),
            'GSE' = mean(GSE_sum), 
            'SD GSE' = sd(GSE_sum),
            'Consequences' = mean(Consequences, na.rm = TRUE),
            'SD Consequences' = sd(Consequences, na.rm = TRUE),
            'Timeline' = mean(Timeline, na.rm = TRUE),
            'SD timeline' = sd(Timeline, na.rm = TRUE),
            'Personal control' = mean(Personal_control, na.rm = TRUE),
            'SD Personal control' = sd(Personal_control, na.rm = TRUE),
            'Treatment control' = mean(Treatment_control, na.rm = TRUE),
            'SD Treatment control' = sd(Treatment_control, na.rm = TRUE),
            'Identity' = mean(Identity, na.rm = TRUE),
            'SD identity' = sd(Identity, na.rm = TRUE),
            'Concern' = mean(Concern, na.rm = TRUE),
            'SD concern' = sd(Concern, na.rm = TRUE),
            'Comprehensibility' = mean(Comprehensibility, na.rm = TRUE),
            'SD Comprehensibility' = sd(Comprehensibility, na.rm = TRUE),
            'Emotions' = mean(Emotions, na.rm = TRUE),
            'SD Emotions' = sd(Emotions, na.rm = TRUE),
            'Brief IPQ' = mean(BIPQ_sum), 
            'SD Brief IPQ' = sd(BIPQ_sum),
            'Calprotectin' = mean(Calprotectin, na.rm = TRUE), 
            'SD Calprotectin' = sd(Calprotectin, na.rm = TRUE),
            'HLQ 1' = mean(HLQ.Scale1), 
            'SD HLQ 1' = sd(HLQ.Scale1),
            'HLQ 2' = mean(HLQ.Scale2), 
            'SD HLQ 2' = sd(HLQ.Scale2),
            'HLQ 3' = mean(HLQ.Scale3), 
            'SD HLQ 3' = sd(HLQ.Scale3),
            'HLQ 4' = mean(HLQ.Scale4), 
            'SD HLQ 4' = sd(HLQ.Scale4),
            'HLQ 5' = mean(HLQ.Scale5), 
            'SD HLQ 5' = sd(HLQ.Scale5),
            'HLQ 6' = mean(HLQ.Scale6), 
            'SD HLQ 6' = sd(HLQ.Scale6),
            'HLQ 7' = mean(HLQ.Scale7), 
            'SD HLQ 7' = sd(HLQ.Scale7),
            'HLQ 8' = mean(HLQ.Scale8), 
            'SD HLQ 8' = sd(HLQ.Scale8),
            'HLQ 9' = mean(HLQ.Scale9), 
            'SD HLQ 9' = sd(HLQ.Scale9),
            'eHLQ 1' = mean(eHLQ.Domain1), 
            'SD eHLQ 1' = sd(eHLQ.Domain1),
            'eHLQ 2' = mean(eHLQ.Domain2), 
            'SD eHLQ 2' = sd(eHLQ.Domain2),
            'eHLQ 3' = mean(eHLQ.Domain3), 
            'SD eHLQ3' = sd(eHLQ.Domain3),
            'eHLQ 4' = mean(eHLQ.Domain4), 
            'SD eHLQ 4' = sd(eHLQ.Domain4),
            'eHLQ 5' = mean(eHLQ.Domain5), 
            'SD eHLQ 5' = sd(eHLQ.Domain5),
            'eHLQ 6' = mean(eHLQ.Domain6), 
            'SD eHLQ 6' = sd(eHLQ.Domain6),
            'eHLQ 7' = mean(eHLQ.Domain7), 
            'SD eHLQ 7' = sd(eHLQ.Domain7))
con.profiles <- as.data.frame(t(con.profiles))
colnames(con.profiles) <- c('Cluster 1', 'Cluster 2', 'Cluster 3')
knitr::kable(con.profiles[-1, ])
```

Cluster profiles for categorical variables 

```{r}
HL.cat <- HL_DHL[, c(1, 3, 7, 9, 13:18, 20:24)]
HL.cat$Cluster <- W2c$Cluster
HL.cat$calpro <- ifelse(is.na(HL_DHL$Calprotectin), NA,
                        ifelse(HL_DHL$Calprotectin >= 250, 1, 0))



# Get sum for each category for each cluster
cat.tbl <- HL.cat %>%
  group_by(Cluster) %>%
  summarise(
    n(),
    "Male" = sum(gender == "Male"),
    "Female" = sum(gender == "Female"),
    "Elementary school" = sum(education == "Elementary school", na.rm = TRUE),
    "Secondary school" = sum(education == "Secondary school", na.rm = TRUE),
    "University college/university ≤ 4 years" = sum(
      education == "University college or university, up to 4 years", na.rm = TRUE),
    "University college/university ≥ 5 years" = sum(
      education == "University college or university, over 5 years", na.rm = TRUE), 
    "Crohn's disease" = sum(diagnose == "Crohn's disease"),
    "Ulcerative colitis" = sum(diagnose == "Ulcerative colitis"),
    "Surgery" = sum(surgery == "Surgery"),
    "Disease activity above threshold" = sum(
      IBD_disease_activity == "Over treshold", na.rm = TRUE),
    "Receiving treatment" = sum(treatment == "Yes"),
    "5-ASA" = sum(ASA_5 == "Yes"),
    "Immunosuppressive" = sum(immunosuppressive == "Yes"),
    "Biological treatment" = sum(biological == "Yes"),
    "Corticosteroids" = sum(corticosteroids == "Yes"),
    "Disease activity(calprotectin)" = sum(calpro == 1, na.rm = T)
  )
# print summaries
knitr::kable(as.data.frame(t(cat.tbl)))
```

## Bivariate analysis of metadata distribution across clusters 

Permuted ANOVA over continuous variables

```{r, message=FALSE}
library(coin)
set.seed(43)

HL.con$Cluster <- as.factor(HL.con$Cluster)
cols.test <- c(1:23, 25:32)
p.values <- numeric(length(cols.test))
for (i in seq_along(cols.test)) {
  var.name <- HL.con[cols.test[i]]
  formula <- as.formula(paste(var.name, "~ Cluster"))
  perm.aov <- oneway_test(formula, data = HL.con, 
                          distribution = approximate(nresample = 9999), 
                          na.rm = TRUE)
  p.values[i] <- pvalue(perm.aov)
}

corrected.p <- p.adjust(p.values, method = "BH")
p.anov <- data.frame(
  Name = colnames(HL.con[, c(1:23, 25:32)]),
  Cor.P = corrected.p
)
knitr::kable(p.anov[corrected.p < .05, ])
```

Post-hoc for ANOVA

```{r}
anov.vars <- colnames(HL.con[, c(3, 6, 9, 10, 23, 25:26, 30, 32)])
clusters <- unique(HL.con$Cluster)
p.vals.T <- list()
set.seed(43)

for (i in anov.vars) {
  var.res <- list()
  var.p <- list()
  for (j in 1:(length(clusters)-1)){
    for (k in (j + 1):length(clusters)){
      group1 <- clusters[j]
      group2 <- clusters[k]
      
      test <- independence_test(as.formula(paste(i, "~ Cluster")),
                                data = subset(HL.con, Cluster %in% c(group1, group2)),
                                distribution = approximate(nresample = 9999))
      name <- paste(group1, "vs.", group2)
      var.p[[name]] <- pvalue(test)
    }
  }
  p.vals.T[[i]] <- var.p
}

p.vals.T <- unlist(p.vals.T)

adj.p.vals <- p.adjust(p.vals.T, method = "BH")

adj.p.vals[adj.p.vals < .05]
```

Chi-square test over nominal variables

```{r, warning=FALSE}
ext.nom <- subset(HL.cat, select = c(gender, diagnose, surgery, 
                                     IBD_disease_activity, treatment, ASA_5,
                                     immunosuppressive, biological, corticosteroids, calpro,
                                     Cluster))

nom.vars <- names(ext.nom)
nom.vars <- nom.vars[-11]

# Create empty lists to store results
test.stat <- list()
p.val <- numeric()

# Designate variables to Chi-square/Fisher test depending on contingency table

for (i in 1:length(nom.vars)) {
   cat.tab <- table(ext.nom[, i], ext.nom$Cluster)
  if (any(cat.tab < 5)) {
    fisher <- fisher.test(cat.tab, workspace = 2e8, hybrid = T)
    test.stat[[i]] <- fisher
    p.val[i] <- fisher$p.value
  } else {
    chisq <- chisq.test(cat.tab)
    test.stat[[i]] <- chisq
    p.val[i] <- chisq$p.value
  }
}

# Give corresponding names in list
names(test.stat) <- nom.vars

# Adjust p-values with Hochberg correction
adj.p <- p.adjust(p.val, method = "BH")


# View significant results and adjusted p-value
p.chi <- data.frame(
  Variables = colnames(ext.nom[, -11]),
  p.vals = adj.p
)

knitr::kable(p.chi[adj.p < .05, ])
```

Post hoc of nominal variables

```{r, message=FALSE}
library(rstatix)
sig.noms <- ext.nom[, c(2, 3, 6, 8, 11)]

# Diagnose
diagnosis.tab <- table(sig.noms$Cluster, sig.noms$diagnose)
dimnames(diagnosis.tab) <- list(
  group = c("Cluster 1", "Cluster 2", "Cluster 3"),
  Diagnosis = c("CD", "UC")
)
pairwise.dia <- pairwise_prop_test(diagnosis.tab)

# Surgery
surg.tab <- table(sig.noms$Cluster, sig.noms$surgery)
dimnames(surg.tab) <- list(
  group = c("Cluster 1", "Cluster 2", "Cluster 3"),
  Surgery = c("No", "Yes")
)
pairwise.surg <- pairwise_prop_test(surg.tab)

# 5-ASA
asa.tab <- table(sig.noms$Cluster, sig.noms$ASA_5)
dimnames(asa.tab) <- list(
  group = c("Cluster 1", "Cluster 2", "Cluster 3"),
  ASA = c("No", "Yes")
)
pairwise.asa <- pairwise_prop_test(asa.tab)

# Biological
bio.tab <- table(sig.noms$Cluster, sig.noms$biological)
dimnames(bio.tab) <- list(
  group = c("Cluster 1", "Cluster 2", "Cluster 3"),
  biological = c("No", "Yes")
)
pairwise.bio <- pairwise_prop_test(bio.tab)

# Collect p-values
p.chis  <- c(pairwise.dia$p, pairwise.surg$p, pairwise.asa$p, pairwise.bio$p)
p.chis.adj <- p.adjust(p.chis, method = "BH")
names(p.chis.adj) <- c("Diagnosis cluster 1 vs. 2", 
                       "Diagnosis cluster 1 vs. 3", 
                       "Diagnosis cluster 2 vs. 3",
                       "Surgery cluster 1 vs. 2", 
                       "Surgery cluster 1 vs. 3", 
                       "Surgery cluster 2 vs. 3",
                       "5-ASA cluster 1 vs. 2", 
                       "5-ASA cluster 1 vs. 3", 
                       "5-ASA cluster 2 vs. 3",
                       "Biological cluster 1 vs. 2", 
                       "Biological cluster 1 vs. 3", 
                       "Biological cluster 2 vs. 3")
print(p.chis.adj[p.chis.adj < .05])
```

Kruskal Wallis test over ordinal variable (education)

```{r}
kruskal.test(Cluster ~ education, data = HL.cat)
```